---
title: "Advanced Data Manipulation and Modelling with `tidyverse`"
subtitle: "Tidy Evaluation, Nested Data, and Model Building"
author: "Professor Alex"
format:
  revealjs:
    monofont: DejaVu Sans Mono
    theme: default
    logo: "unilogo_noir_300dpi.png"
    footer: "Advanced R Programming and Modelling - Lecture 7"
    slide-number: true
    incremental: false
    preview-links: true
    code-line-numbers: false
    highlight-style: printing
    code-block-height: 500px
    code-copy: true
    code-link: true
    link-external-icon: false
    link-external-newwindow: true
---

```{r setup}
#| echo: false
#| eval: true
#| message: false
# Load necessary libraries
library(conflicted)
conflict_prefer("filter", "dplyr")
library(tidyverse)
library(knitr)
library(kableExtra)
library(broom)
```

# Introduction to Tidy Evaluation in `dplyr` {background-color="#43464B"}

## Overview

- Modern data manipulation in R with `tidyverse`
- Focus on tidy evaluation principles
- Practical applications in data science workflows

## What We'll Cover Today

:::: {.columns}

::: {.column width="50%"}
**Part 1: Tidy Evaluation in `dplyr`**

- Data masking
- Tidy selection
- Programming with `dplyr`
:::

::: {.column width="50%"}
**Part 2: Nested Data & Modelling**

- Nested data frames
- Model building
- Result aggregation
:::

::::

# Part 1: Tidy Evaluation in `dplyr` {background-color="#43464B"}

## What is Tidy Evaluation?
- **Tidy Evaluation** is a form of non-standard evaluation used in the `tidyverse`.
- It allows simpler expressions for data manipulation, making code more concise and readable.
- Two main concepts: **Data Masking** and **Tidy Selection**.

## Why Use Tidy Evaluation?
- Reduces the need for `df$column` syntax; instead, use `column` directly.
- Enables smooth data exploration and efficient manipulation, especially in interactive analysis.
- Supports consistent and flexible workflows across multiple `dplyr` functions.

## Practical Example: Tidy Evaluation {.scrollable}

```{r}
#| echo: true
#| eval: false
# Filtering without Tidy Evaluation
starwars[starwars$homeworld == "Naboo" & starwars$species == "Human", ]

# Filtering with Tidy Evaluation
starwars %>% dplyr::filter(homeworld == "Naboo", species == "Human")
```

1. Use the `$` operator to reference columns.
    - Not recommended for `dplyr` workflows.
    - Good when developing scripts or functions to avoid external dependencies.
2. Use the `dplyr::filter()` function with direct column references.
    - Cleaner and more readable.
    - Preferred method for `dplyr` workflows.

## Overview of Data Masking

- **Data Masking** allows using data frame columns as if they were variables in the environment.
- Avoids repetitive references to the data frame.
- Key for efficient filtering, summarizing, and mutating.

---

## Data Masking in Action
- Distinguishes between **env-variables** (programming variables) and **data-variables** (statistical variables).
- Env-variables are created with `<-`, while data-variables are within the data frame.

---

## Practical Example: Data Masking

::: {.panel-tabset}

### Env-Variables

```{r}
#| echo: true
# Using df as an env-variable
df <- tibble::tibble(x = runif(3), y = runif(3))
df %>% dplyr::mutate(z = x + y)
```

### Data-Variables

```{r}
#| echo: true
# Filtering with Data Masking
# Here, "homeworld" and "species" are data-variables
starwars %>%
  dplyr::filter(homeworld == "Naboo" & species == "Human") %>% 
  dplyr::select(name, homeworld, species)
```

:::

## Tidy Selection

**What is Tidy Selection?**

- **Tidy Selection** is a set of tools provided by `tidyselect` package to easily select columns by name, type, or position.
- Allows for selecting columns dynamically within functions like
    - `dplyr::select()`
    - `dplyr::mutate()`
    - `dplyr::across()`

---

## Tidy Selection Syntax

- Examples: 
    - `tidyselect::starts_with()`
    - `tidyselect::ends_with()`
    - `tidyselect::where()`
- Can specify columns using multiple criteria within a single function.

---

## Practical Example: Tidy Selection

::: {.panel-tabset}

### Prefix

```{r}
#| echo: true
#| eval: true

# Selecting columns that start with "c"
mtcars %>% select(starts_with("c"))
```

### Contains

```{r}
#| echo: true
#| eval: true

# Selecting columns that contain a letter "t"
mtcars %>% select(contains("t"))
```

### Match

```{r}
#| echo: true
#| eval: true

# Selecting columns that start with letters a-g
mtcars %>% select(matches("^[a-g]"))
```

### Type

```{r}
#| echo: true
#| eval: true

# Selecting all numeric columns
mtcars %>% select(where(is.numeric))
```

:::

## Tidy Selection: Common Selectors

1. Select variables that match a pattern
    - `starts_with()`, `ends_with()`, `contains()`, `matches()`, `num_range()`
2. Select variables from character vector
    - `all_of()`, `any_of()`
3. Select all variables or the last variable
    - `everything()`, `last_col()`
4. Select variables with a function
    - `where()`

## Programming with `dplyr`

**What is Indirection in `dplyr`?**

- **Indirection** refers to using variables dynamically within `dplyr` functions.
- Useful in functions and loops, especially when column names are stored as strings or variables.

---

## Techniques for Indirection
- Use `{{{ }}}` to embrace expressions within function arguments.
- `.data[[ ]]` helps when column names are in character vectors.

---

## Practical Example: Indirection with Embracing

```{r}
#| echo: true
#| eval: true
# Embracing with {{ }}
var_summary <- function(data, var) {
  data %>% dplyr::summarise(
    n = dplyr::n(),
    min = min({{ var }}),
    max = max({{ var }})
  )
}
mtcars %>%
  dplyr::group_by(cyl) %>%
  var_summary(mpg)
```

## Across Multiple Columns {.smaller}

- `dplyr::across()` function allows applying operations across **multiple columns**.
- Useful for summarizing, mutating, and filtering multiple columns.

**Syntax:**

> `across(.cols, .fns, .names = NULL, .unpack = FALSE)`

- `.cols`: `<tidy-select>` Columns to apply the function.
- `.fns`: Function to apply to the columns.
    - A function, e.g. `mean`
    - A `purrr`-style lambda function, e.g. `~mean(.x, na.rm = TRUE)`
    - A named list of functions, e.g. `list(mean = mean, n_miss = ~ sum(is.na(.x)))`

## Practical Example: Using `across()`

::: {.panel-tabset}

### Single Function

```{r}
#| echo: true
#| eval: true

# Applying mean across multiple columns
vars <- c("mpg", "vs")
mtcars %>%
  dplyr::summarise(dplyr::across(
    tidyselect::all_of(vars),
    mean
  ))
```

### Multiple Functions

```{r}
#| echo: true
#| eval: true

# Applying multiple functions across numeric columns
mtcars %>%
  dplyr::summarise(dplyr::across(
    tidyselect::where(is.numeric), 
    list(avg = mean, std = sd)
  ))
```

### String to Factor

```{r}
#| echo: true
#| eval: true

# Converting character columns to factors
starwars %>%
  mutate(across(where(is.character), as.factor)) %>% 
  select(name, where(is.factor))
```

:::

## Flexible Summary Functions

- Functions can accept variable column expressions to compute summaries dynamically.
- `dplyr::across()` is key for creating flexible summaries over multiple columns.

**Custom Summarization Syntax**

- Use `dplyr::across()` inside `dplyr::summarise()` to calculate statistics on chosen columns.
- Control output column names using `.names` within `dplyr::across()`.

---

## Practical Example: Custom Summarization {.scrollable}

::: {.panel-tabset}

### Embracing `{{{ }}}`

```{r}
#| echo: true
#| eval: true
# Function to calculate means of numeric columns
summarise_mean <- function(data, vars) {
  data %>% 
    summarise(
      n = n(),
      across({{ vars }}, ~ mean(.x, na.rm = TRUE), .names = "{.col}_avg")
    )
}
mtcars %>% group_by(cyl) %>% summarise_mean(where(is.numeric))
```

### Using `...`

```{r}
#| echo: true
#| eval: true
# Function to calculate means of numeric columns
summarise_mean <- function(.data, ...) {
  .data %>% 
    summarise(
      n = n(),
      across(c(...), ~ mean(.x, na.rm = TRUE), .names = "{.col}_avg")
    )
}
mtcars %>% group_by(cyl) %>% summarise_mean(mpg, disp, hp, wt)
```

:::

## Name Injection in Variable Names

**Dynamic Name Creation with Name Injection**

- `:=` allows creating dynamic names for variables in a pipeline.
- Useful when column names are generated programmatically or from environment variables.

---

## Practical Example: Name Injection

:::: {.columns}

::: {.column width="38%"}

```{r}
#| echo: true
#| eval: true

# Dynamic name assignment
name <- "dynamic_var"
tibble("{name}" := 2)
```

:::

::: {.column width="58%"}

```{r}
#| echo: true
#| eval: true

# Using dynamic name within a function
my_df <- function(x) {
  tibble("{{ x }}_2" := x * 2)
}
my_var <- 10
my_df(my_var)
```

:::

::::

# Part 2: Nested Data & Modelling {background-color="#43464B"}

## What is a Nested Data Frame?
- A **Nested Data Frame** contains one or more columns as lists of data frames.
- Useful for organizing grouped data or subsetting for complex transformations and modeling.

## Creating Nested Data Frames
- Use `tidyr::nest()` to group columns into list-columns.
- Each list entry represents a data subset, allowing efficient grouping and nested operations.

```{r}
#| echo: true
#| eval: true
#| output-location: column

# Creating a nested data frame by grouping
df <- tribble(
  ~g, ~x, ~y,
   1,  1,  2,
   2,  4,  6,
   2,  5,  7,
   3, 10,  NA
)
df %>% tidyr::nest(data = c(x, y))
```

## Practical Example: (Un)Nesting

::: {.panel-tabset}

### Original Data

```{r}
#| echo: true
#| eval: true

# Display some columns
starwars %>% select(species, name, height, mass, gender, homeworld)
```

### Nested Data

```{r}
#| echo: true
#| eval: true
#| output-location: column

# Nesting data by species
starwars %>%
  dplyr::group_by(species) %>%
  tidyr::nest()
```

:::

## Using Nested Data for Modeling
- Nested data is ideal for **fitting models across groups**.
- Each subset in a list-column can hold an individual model or analysis result.

---

## Practical Example: Nesting `mtcars` for Modeling
- Group `mtcars` by `cyl` (number of cylinders) and nest data within each group.
```{r}
#| echo: true
#| eval: true
#| output-location: column
# Nesting mtcars data by cylinder
mtcars_nested <- mtcars %>%
  group_by(cyl) %>%
  nest()
mtcars_nested
```

---

## Building Models on Nested Data
- Use `purrr::map()` to fit models within each nested group.
- Store each model in a new list-column.

```{r}
#| echo: true
#| eval: true

# Fit models to each subset
mtcars_nested <- mtcars_nested %>% 
  mutate(model = map(data, \(df) lm(mpg ~ wt, data = df)))
mtcars_nested
```

---

## Extracting Predictions
- Generate predictions for each model using `map()` to apply `predict()`.

```{r}
#| echo: true
#| eval: true
# Generate predictions
mtcars_nested <- mtcars_nested %>%
  mutate(predictions = map(model, predict))
mtcars_nested
```

## Applying `broom` for Model Summaries {.smaller}

:::: {.columns}

::: {.column width="50%"}
**Available Functions:**

- `broom::tidy()`
- `broom::glance()`
- `broom::augment()`
:::

::: {.column width="50%"}
**Common Use Cases:**

- Model coefficients (coefficients, p-values)
- Model statistics (R-squared, AIC)
- Predictions and residuals
:::

::::

- `broom` provides functions like `tidy()`, `augment()`, and `glance()` to convert model outputs to data frames.
- This allows for consistent analysis and visualization of multiple models' outputs.
- For linear models, `lm()`, see:
  - `broom::tidy.lm()`
  - `broom::glance.lm()`
  - `broom::augment.lm()`

---

## Practical Example: Tidying with `broom::tidy()` {.smaller}
```{r}
#| echo: true
#| eval: true
#| output-location: fragment
# Tidying model results
library(broom)
mtcars_nested <- mtcars_nested %>%
  mutate(tidy_results = purrr::map(model, broom::tidy))

# Unnesting to view all results in a flat format
mtcars_nested %>% unnest(tidy_results)
```

---

## Advanced Tidying: `augment()` and `glance()` {.scrollable}

- **`augment()`**: Adds model predictions and residuals to the original data.
- **`glance()`**: Summarizes model fit statistics.

```{r}
#| echo: true
#| eval: true

# Adding model summaries
mtcars_nested <- mtcars_nested %>%
  mutate(
    augmented = map(model, augment),
    glance_results = map(model, glance)
  )
```

---

## Model Fit Summaries {.smaller}

::: {.panel-tabset}

### Data

```{r}
#| echo: true
#| eval: true
#| code-overflow: scroll

mtcars_nested %>% 
  select(cyl, data, model, tidy_results, augmented, glance_results)
```

### Tidy

```{r}
#| echo: true
#| eval: true
#| code-overflow: scroll

mtcars_nested %>%
  select(cyl, data, model, tidy_results) %>%
  unnest(tidy_results)
```

### Augmented

```{r}
#| echo: true
#| eval: true
#| code-overflow: scroll

mtcars_nested %>%
  select(cyl, data, model, augmented) %>%
  unnest(augmented)
```

### Glance

```{r}
#| echo: true
#| eval: true
#| code-overflow: scroll

mtcars_nested %>%
  select(cyl, data, model, glance_results) %>%
  unnest(glance_results)
```

:::

---

## Practical Application: Model Comparison {.smaller}

- Easily compare models across groups with tidied outputs.
- Sort or filter by fit statistics, visualize parameter estimates, or compare residuals.

```{r}
#| echo: true
#| eval: true

# Example: Sorting by R-squared from glance results
mtcars_nested %>%
  select(cyl, data, model, glance_results) %>%
  unnest(glance_results) %>%
  arrange(desc(r.squared))
```

# Model Selection: Comparing Different Models {background-color="#43464B"}

## Linear Regression Models

- Create multiple linear models on `mtcars` data for each cylinder group.
- Fit models and store results in a nested data frame.
- Nest the `mtcars` data by cylinder group:

```{r}
#| echo: true
#| eval: true

mtcars_nested <- mtcars %>%
  group_by(cyl) %>%
  nest()
mtcars_nested
```

## Fit Linear Models {.smaller}

- Consider building linear models for each cylinder group:
  - `mpg ~ wt`
  - `mpg ~ wt + hp`
  - `mpg ~ wt + hp + qsec`
  
```{r}
#| echo: true
#| eval: true

# Fit linear models
mtcars_lm <- mtcars_nested %>% 
  mutate(
    model_wt = map(data, \(df) lm(mpg ~ wt, data = df)),
    model_wt_hp = map(data, \(df) lm(mpg ~ wt + hp, data = df)),
    model_wt_hp_qsec = map(data, \(df) lm(mpg ~ wt + hp + qsec, data = df))
  )
mtcars_lm
```

## All Model Statistics {.smaller}

- Use `pivot_longer()` to reshape the data for comparison.
- Using `broom::glance()` to extract model statistics.

```{r}
#| echo: true
#| eval: true

# Extract model statistics
mtcars_glance <- mtcars_lm %>%
  pivot_longer(
    cols = starts_with("model"),
    names_to = "model",
    values_to = "model_fit"
  ) %>% 
  mutate(
    model_stats = map(model_fit, glance)
  )
mtcars_glance
```

## Unnest and Compare Models {.smaller}

- Use `unnest()` to extract model statistics for comparison.

```{r}
#| echo: true
#| eval: true

# Pivot longer for comparison
mtcars_lm_stats <- mtcars_glance %>%
  unnest(model_stats)
mtcars_lm_stats
```

## Model Selection  {.smaller}

- Select the best model based on, for example, lowest BIC values.
- Use `dplyr::slice_min()` to select the row with the minimum BIC value.

```{r}
#| echo: true
#| eval: true

# Select best model by BIC
mtcars_lm_stats %>%
  slice_min(BIC) %>% 
  relocate(BIC, .after = model_fit)
```

## Summary and Best Practices

- Use tidy evaluation for cleaner code
- Leverage `nest`'ed data frames for grouped analyses
- Apply `broom` for consistent model outputs
- Consider computational efficiency and readability

## Resources

:::: {.columns}

::: {.column width="50%"}
**Documentation**

- [tidyselect: Selection Helpers](https://tidyselect.r-lib.org/reference/language.html)
- [Nested data](https://tidyr.tidyverse.org/articles/nest.html)
- [dplyr programming](https://dplyr.tidyverse.org/articles/programming.html)
- [broom documentation](https://broom.tidymodels.org)
- [broom and dplyr](https://broom.tidymodels.org/articles/broom_and_dplyr.html)
:::

::: {.column width="50%"}
**Further Reading**

- R for Data Science
- Advanced R by Hadley Wickham
- tidyverse blog posts
:::

::::
